% !TEX root = ../Monografia.tex
% ---------------------------------------------------------
% BIBLIOGRAFIA
% ---------------------------------------------------------
\begin{thebibliography}{99}

% ---------------------------------------------------------
% TRABALHOS ACADÊMICOS RELACIONADOS
% ---------------------------------------------------------

\bibitem{pathak2012energy}
PATHAK, A.; HU, Y. C.; ZHANG, M.
\newblock \textit{Fine-grained Power Modeling for Smartphones Using System Call Tracing}.
\newblock In: \emph{Proceedings of the European Conference on Computer Systems}.
\newblock ACM, 2012.

\bibitem{liu2019profiling}
LIU, Y.; ZHOU, X.; ZHANG, H.
\newblock \textit{Profiling and Diagnosing Performance Issues in Mobile Applications}.
\newblock \emph{IEEE Software}, v. 36, n. 5, p. 80–87, 2019.

\bibitem{he2019crossplatform}
HE, J.; LI, S.; WANG, Y.
\newblock \textit{Performance Comparison of Cross-Platform Mobile Development Frameworks}.
\newblock \emph{Journal of Systems and Software}, v. 152, p. 1–16, 2019.

\bibitem{banerjee2020automated}
BANERJEE, S.; ROY, S.; DUTTA, K.
\newblock \textit{Automated Runtime Monitoring for Mobile Applications}.
\newblock \emph{ACM Transactions on Software Engineering and Methodology}, v. 29, n. 4, 2020.

\bibitem{hao2013estimating}
HAO, S.; LI, D.; HALFOND, W.
\newblock \textit{Estimating Mobile Application Energy Consumption Using Program Analysis}.
\newblock In: \emph{Proceedings of the International Conference on Software Engineering (ICSE)}.
\newblock IEEE, 2013.

\bibitem{li2014empirical}
LI, D.; HALFOND, W.
\newblock \textit{An Empirical Study of Energy Consumption of Android Applications}.
\newblock In: \emph{International Conference on Software Maintenance and Evolution}.
\newblock IEEE, 2014.

% ---------------------------------------------------------
% TRABALHOS ACADÊMICOS (TCC / DISSERTAÇÕES)
% ---------------------------------------------------------

\bibitem{silva2022mobile}
SILVA, R. A.
\newblock \textit{Análise de Consumo Energético em Aplicações Móveis Android}.
\newblock Dissertação (Mestrado em Ciência da Computação) – Universidade Federal de Pernambuco, 2022.

\bibitem{santos2023flutter}
SANTOS, L. F.
\newblock \textit{Avaliação de Desempenho em Aplicações Flutter: Um Estudo Experimental}.
\newblock Trabalho de Conclusão de Curso – Universidade de São Paulo, 2023.

\bibitem{rodrigues2021mobile}
RODRIGUES, M. P.
\newblock \textit{Ferramentas de Monitoramento de Desempenho em Aplicações Móveis}.
\newblock Trabalho de Conclusão de Curso – Universidade Federal de Minas Gerais, 2021.

% ---------------------------------------------------------
% DOCUMENTAÇÃO DE FERRAMENTAS
% ---------------------------------------------------------

\bibitem{android2024}
ANDROID DEVELOPERS.
\newblock \textbf{Android Profiler}.
\newblock Disponível em: \url{https://developer.android.com/studio/profile}.
\newblock Acesso em: 05 out. 2025.

\bibitem{aosp2024}
ANDROID OPEN SOURCE PROJECT.
\newblock \textbf{Perfetto Tracing}.
\newblock Disponível em: \url{https://perfetto.dev}.
\newblock Acesso em: 05 out. 2025.

\bibitem{apple2024}
APPLE.
\newblock \textbf{Xcode Instruments}.
\newblock Disponível em: \url{https://developer.apple.com/xcode/instruments/}.
\newblock Acesso em: 05 out. 2025.

\bibitem{meta2024}
META.
\newblock \textbf{Flipper – React Native Debugging Tool}.
\newblock Disponível em: \url{https://fbflipper.com}.
\newblock Acesso em: 05 out. 2025.

\bibitem{google2024devtools}
GOOGLE.
\newblock \textbf{Flutter DevTools}.
\newblock Disponível em: \url{https://docs.flutter.dev/development/tools/devtools/overview}.
\newblock Acesso em: 05 out. 2025.

\bibitem{google2023flutter}
GOOGLE.
\newblock \textbf{Flutter: Build Apps for Any Screen}.
\newblock Disponível em: \url{https://flutter.dev}.
\newblock Acesso em: 22 ago. 2025.

\bibitem{meyer2022performance}
MEYER, J.; KRAUSE, A.
\newblock \textit{Performance Evaluation of Flutter vs React Native in Production Environments}.
\newblock \emph{International Journal of Mobile Computing and Multimedia Communications}, v. 8, n. 2, p. 112--125, 2022.

\bibitem{dart2023}
DART TEAM.
\newblock \textbf{Dart: A Client-Optimized Language for Fast Apps on Any Platform}.
\newblock Disponível em: \url{https://dart.dev}.
\newblock Acesso em: 22 ago. 2025.

% ---------------------------------------------------------
% ARTIGOS SOBRE ENERGIA E DESEMPENHO
% ---------------------------------------------------------

\bibitem{cheng2020energy}
CHENG, L.; WANG, X.
\newblock \textit{Energy Consumption Analysis in Mobile Applications: A Framework-based Approach}.
\newblock \emph{Journal of Software Engineering}, 15(4), 45–59, 2020.

\bibitem{zhang2021energy}
ZHANG, Y.; LIU, J.; CHEN, F.
\newblock \textit{Evaluating Energy Efficiency of Flutter and React Native Applications}.
\newblock In: \emph{Proceedings of the International Conference on Mobile Systems}.
\newblock ACM, 2021.

\bibitem{johnson2020monitoring}
JOHNSON, R.
\newblock \textit{Monitoring System Calls and Frame Rendering Events for Energy Optimization}.
\newblock \emph{Software: Practice and Experience}, 50(9), 1652–1668, 2020.

\bibitem{kumar2019optimization}
KUMAR, P.; GUPTA, R.
\newblock \textit{Optimizing Resource Consumption in Mobile Apps: Techniques and Tools}.
\newblock \emph{International Journal of Software Engineering}, 12(4), 234–250, 2019.

% ---------------------------------------------------------
% SUSTENTABILIDADE
% ---------------------------------------------------------

\bibitem{google2024sustainability}
GOOGLE SUSTAINABILITY.
\newblock \textbf{Reducing Energy Consumption in Software}.
\newblock Disponível em: \url{https://sustainability.google}.
\newblock Acesso em: 05 out. 2025.

\bibitem{ieee2023sustainability}
IEEE.
\newblock \textbf{Sustainability in Software Engineering}.
\newblock Disponível em: \url{https://www.ieee.org/sustainability}.
\newblock Acesso em: 05 out. 2025.

\bibitem{unep2023}
UNITED NATIONS ENVIRONMENT PROGRAMME (UNEP).
\newblock \textbf{Digital Sustainability Initiatives}.
\newblock Disponível em: \url{https://www.unep.org/resources}.
\newblock Acesso em: 05 out. 2025.

% ---------------------------------------------------------
% METODOLOGIA CIENTÍFICA
% ---------------------------------------------------------

\bibitem{gil2017}
GIL, A. C.
\newblock \textbf{Métodos e Técnicas de Pesquisa Social}.
\newblock 7. ed. São Paulo: Atlas, 2017.

\bibitem{lakatos2021}
LAKATOS, E. M.; MARCONI, M. A.
\newblock \textbf{Fundamentos de Metodologia Científica}.
\newblock 9. ed. São Paulo: Atlas, 2021.

\bibitem{yin2015}
YIN, R. K.
\newblock \textbf{Estudo de Caso: Planejamento e Métodos}.
\newblock 5. ed. Porto Alegre: Bookman, 2015.

\bibitem{wohlin2012}
WOHLIN, C. et al.
\newblock \textbf{Experimentation in Software Engineering}.
\newblock Springer, 2012.

% ---------------------------------------------------------
% ENGENHARIA DE SOFTWARE
% ---------------------------------------------------------

\bibitem{pressman2020engenharia}
PRESSMAN, R. S.; MAXIM, B. R.
\newblock \textbf{Engenharia de Software: Uma Abordagem Profissional}.
\newblock 9. ed. Porto Alegre: AMGH, 2020.

\bibitem{kitchenham2007guidelines}
KITCHENHAM, B.; CHARTERS, S.
\newblock \textbf{Guidelines for Performing Systematic Literature Reviews in Software Engineering}.
\newblock EBSE Technical Report, Keele University, 2007.

\bibitem{sommerville2019software}
SOMMERVILLE, I.
\newblock \textbf{Engenharia de Software}.
\newblock 10. ed. São Paulo: Pearson, 2019.

\bibitem{martin2018clean}
MARTIN, R. C.
\newblock \textbf{Clean Architecture: A Craftsman’s Guide to Software Structure and Design}.
\newblock Prentice Hall, 2018.

\bibitem{evans2004domain}
EVANS, E.
\newblock \textbf{Domain-Driven Design: Tackling Complexity in the Heart of Software}.
\newblock Addison-Wesley, 2004.

% ---------------------------------------------------------
% REFERÊNCIAS ADICIONADAS NA REVISÃO (verificar antes da defesa)
% Fontes identificadas em IEEE Xplore, ACM DL, arXiv e Springer.
% ---------------------------------------------------------

\bibitem{ravindranath2012appinsight}
RAVINDRANATH, L.; PADHYE, J.; AGARWAL, S.; MAHAJAN, R.; OBERMILLER, I.; SHAYANDEH, S.
\newblock \textit{AppInsight: Mobile App Performance Monitoring in the Wild}.
\newblock In: \emph{Proceedings of the 10th USENIX Symposium on Operating Systems Design and Implementation (OSDI 12)}.
\newblock USENIX Association, 2012. p. 107--120.

\bibitem{verma2023largescale}
VERMA, A.; TRIVEDI, A.; PANDEY, H. M.
\newblock \textit{A large-scale empirical study on mobile performance: energy, run-time and memory}.
\newblock \emph{Empirical Software Engineering}, v. 28, art. 161, 2023.
\newblock DOI: 10.1007/s10664-023-10391-y.

\bibitem{horn2023nativeweb}
HORN, R. et al.
\newblock \textit{Native vs Web Apps: Comparing the Energy Consumption and Performance of Android Apps and their Web Counterparts}.
\newblock In: \emph{Proceedings of the 10th International Conference on Mobile Software Engineering and Systems (MOBILESoft)}.
\newblock ACM, 2023.

\bibitem{dornauer2023energysaving}
DORNAUER, B.; FELDERER, M.
\newblock \textit{Energy-Saving Strategies for Mobile Web Apps and their Measurement: Results from a Decade of Research}.
\newblock In: \emph{Proceedings of the 10th International Conference on Mobile Software Engineering and Systems (MOBILESoft)}.
\newblock ACM, 2023.

\bibitem{sun2023energyinefficiency}
SUN, Y. et al.
\newblock \textit{Energy inefficiency diagnosis for Android applications: a literature review}.
\newblock \emph{Frontiers of Computer Science}, v. 17, n. 1, art. 171201, 2023.
\newblock DOI: 10.1007/s11704-021-0532-4.

\bibitem{toni2023apienergy}
TONI, C. et al.
\newblock \textit{Energy-Consumption Estimation of API-usage in Smartphone Apps via Static Analysis}.
\newblock In: \emph{Proceedings of the 20th International Conference on Mining Software Repositories (MSR)}.
\newblock IEEE, 2023.
\newblock DOI: 10.1109/MSR59073.2023.00057.

\end{thebibliography}